Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-15 paper_io_sim.tex
Edits in results and conclusion

        \wrench simulation errors are substantially lower with 100~GB
        files than with 20~GB files, due to the fact that part of the
        100~GB file needs to be read and written to disk, the only storage
        device in \wrench, as it does not fit in cache. Conversely,
        simulation errors of the Python prototype and \wrench-cache are higher with
        100~GB files than with 20~GB files, due to idiosyncrasies in the kernel
        flushing and eviction strategies.

        Simulated memory profiles are highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). Used memory consists of both
        cached and anonymous memory that cannot be evicted at the current
        time. With 20~GB files, memory profiles almost exactly match the
        real ones, although dirty data seems to be flushing faster in real
        life than in simulation \tristan{any idea why?}, a behavior also
        observed with 100~GB files. With 100~GB files, used memory reaches
        total memory during the first write, triggering dirty data
        flushing. Used memory drops to cached memory when application tasks
        release anonymous memory. Simulated cached memory is highly
        consistent with real values, except toward the end of Read 3 where
        it slightly increases in simulation but not in reality. This
        occurred due to the fact that after Write 2, File 3 was only partly
        cached in simulation whereas it was entirely cached in the real
        system. In all cases, dirty data remains under the dirty ratio as
        expected. The Python prototype and \wrench-cache exhibit nearly
        identical memory profiles, which reinforces the confidence in our
        implementation.

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, 
        only a part of it is cached. 
        This is due to the fact that the in simulation, data is flushed and evicted 
        from the inactive list first, then from the active list \tristan{whereas in the real execution ...?}. Thus, File 2, which is 
        accessed again by Read 2 and is cached in the active list, is not flushed and evicted, 
        while File 3 is flushed and evicted since it is accessed once only and cached 
        in the inactive list.

        % We conclude from this experiment that our model can accurately
        % simulate cached I/Os in a sequential context. The next experiment
        % evaluates our simulation with concurrent I/Os.

        \subsection{Concurrent application (local storage)}

@@

@@

            \begin{figure*}
            \begin{subfigure}{\linewidth}
                \centering
                \includegraphics[width=\linewidth]{result/multi/figures/multi_nfs.pdf}
            \end{subfigure}
            \caption{NFS results with 3~GB files (averages on 5 repetitions)}
            \label{fig:multi_nfs}
            \end{figure*}

            Page cache simulation importantly reduces the simulation error
            on remote storage as well (Figure~\ref{fig:multi_nfs}). This
            manifests only for reads, as only writethrough cache was
            enabled in the NFS server. Both \wrench and \wrench-cache
            underestimate write times due to the discrepancy between
            simulated and real bandwidths mentioned previously. For reads,
            this discrepancy only impacts the results beyond 22
            applications since before this threshold, most reads result in cache
            hits.
            % \tristan{Dzung, pls check that.}.
            % Yes, disk reads appear from 23 apps indeed.

@@

    Page cache simulation can be instrumental in a number of studies. For
    instance, it is now common for HPC clusters to run applications in
    Linux control groups (cgroups), where resource consumption is limited,
    including memory and therefore page cache usage. Using our simulator,
    it would be possible to study the interaction between memory allocation
    and I/O performance, for instance to improve scheduling algorithms or
    avoid page cache starvation~\cite{zhuang2017}. Our simulator could also
    be utilized to evaluate solutions that reduce the impact of network
    file transfers on distributed applications, such as burst
    buffers~\cite{ferreiradasilva-fgcs-bb-2019}, hierarchical file
    systems~\cite{islam2015triple}, active storage~\cite{5496981}, or
    specific hardware architectures~\cite{hayot2020performance}.

    However, I/Os remain much more complex than is currently available in
    simulation models, which could drastically limit the accuracy of simulations.
    Relevant extensions to this work could include more
    accurate descriptions of anonymous memory usage by applications. 
    With the page cache, I/O times are indeed strongly
    dependent on the amount of available memory, which is itself impacted
    by the use of anonymous memory. File access patterns, 
    including random I/O, might also be worth including in application models,
    as they directly affect page cache content.
    % readahead and persistent storage could also be adeed
    % Simulation results could be made even more accurate by a deeper
    % investigation of page cache flushing and eviction order.

>>>>

        \wrench simulation errors are substantially lower with 100~GB
        files than with 20~GB files, due to the fact that part of the
        100~GB file needs to be read and written to disk, the only storage
        device in \wrench, as it does not fit in cache. Conversely,
        simulation errors of the Python prototype and \wrench-cache are higher with
        100~GB files than with 20~GB files, due to idiosyncrasies in the kernel
        flushing and eviction strategies that could not be easily modeled.

        Simulated memory profiles are highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). Used memory consists of both
        cached and anonymous memory that cannot be evicted at the current
        time. With 20~GB files, memory profiles almost exactly match the
        real ones, although dirty data seems to be flushing faster in real
        life than in simulation \tristan{any idea why?}, a behavior also
        observed with 100~GB files. With 100~GB files, used memory reaches
        total memory during the first write, triggering dirty data
        flushing, and drops back to cached memory when application tasks
        release anonymous memory. Simulated cached memory is highly
        consistent with real values, except toward the end of Read 3 where
        it slightly increases in simulation but not in reality. This
        occurred due to the fact that after Write 2, File 3 was only partially
        cached in simulation whereas it was entirely cached in the real
        system. In all cases, dirty data remains under the dirty ratio as
        expected. The Python prototype and \wrench-cache exhibit nearly
        identical memory profiles, which reinforces the confidence in our
        implementations.

@@

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, 
        only a part of it is cached. 
        This is due to the fact that the Linux kernel tend to not evict 
        pages that belong to files being currently written (File 3 in this case), which we 
        could not easily reproduce in our model.
        \begin{figure*}
            \begin{subfigure}{\linewidth}
                \centering
                \includegraphics[width=\linewidth]{result/multi/figures/multi_nfs.pdf}
            \end{subfigure}
            \caption{NFS results with 3~GB files (averages on 5 repetitions)}
            \label{fig:multi_nfs}
            \end{figure*}

        \subsection{Concurrent applications}

@@

            Page cache simulation importantly reduces the simulation error
            on remote storage as well (Figure~\ref{fig:multi_nfs}). This
            manifests only for reads, as the NFS server used writethrough rather than writeback cache.
            Both \wrench and \wrench-cache
            underestimate write times due to the discrepancy between
            simulated and real bandwidths mentioned previously. For reads,
            this discrepancy only impacts the results beyond 22
            applications since before this threshold, most reads result in cache
            hits.
            % \tristan{Dzung, pls check that.}.
            % Yes, disk reads appear from 23 apps indeed.

@@

    Page cache simulation can be instrumental in a number of studies. For
    instance, it is now common for HPC clusters to run applications in
    Linux control groups (cgroups), where resource consumption is limited,
    including memory and therefore page cache usage. Using our simulator,
    it would be possible to study the interaction between memory allocation
    and I/O performance, for instance to improve scheduling algorithms or
    avoid page cache starvation~\cite{zhuang2017}. Our simulator could also
    be leveraged to evaluate solutions that reduce the impact of network
    file transfers on distributed applications, such as burst
    buffers~\cite{ferreiradasilva-fgcs-bb-2019}, hierarchical file
    systems~\cite{islam2015triple}, active storage~\cite{5496981}, or
    specific hardware architectures~\cite{hayot2020performance}.

    I/Os remain much more complex than is currently available in
    simulation models, which could substantially limit the accuracy of simulations.
    Relevant extensions to this work could include more
    accurate descriptions of anonymous memory usage in applications, 
    which strongly affects I/O times through writeback cache. File access patterns, 
    including random I/O, might also be worth including in application models,
    as they directly affect page cache content.
    % readahead and persistent storage could also be adeed
    % Simulation results could be made even more accurate by a deeper
    % investigation of page cache flushing and eviction order.
